---
title: "Findings Code"
author: "Hyesop Shin"
toc: true
toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
format: html
editor_options: 
  chunk_output_type: console
---

## Intro

This documentation is to demonstrate the procedure of our analysis for Glasgow Low Emission Zone (LEZ). To begin with, we call files that are in `.parquet` `.RData`. We then explore the traffic and NO~2~ data to understand the distribution in two different areas, namely Hope St (city centre) and High St (Edge of LEZ).

Here we use a number of libraries.

```{r}
options(scipen = 999)

library(arrow)
library(data.table)
library(tidyverse)
library(sf)
library(mapview)
library(openair)
library(rstatix)
library(ggpubr)

```


Next part is to import the traffic data. *Note we will execute the NO~2~ on a separate section.*

```{r}
##-SCOOT DATA-##
df22 <- read_parquet("glasgow2022.parquet") ## SCOOT Data
df23 <- read_parquet("glasgow2023.parquet") ## SCOOT Data
monitor <- read_sf("Mointor shp/useful_detector_530.shp") ## SCOOT Monitor
lez_shp <- read_sf("Low_Emission_Zone(LEZ)/Low_Emission_Zone_(LEZ).shp") ## LEZ Boundary

## Additional input
load("SCOOT_SEP22_SEP23.RData")

```

```{r dataclean}

## SCOOT 2023
df23 %>%  ## <- df is the raw SCOOT Data
  as_tibble() %>% 
  mutate(timestamp = lubridate::as_datetime(timeStamp),
         dt_date = date(timestamp),
         dt_month = month(timestamp),
         dt_weekdays = weekdays(timestamp),
         dt_year = year(timestamp)) |> 
  #filter(between(dt_month, 6, 8)) |> 
  filter(dt_month >= 8) |> 
  select(siteId, dt_date, dt_year, dt_month, dt_weekdays, flow) -> df23_dt


# Adding September 2023
dt23_sep |> 
  select(-year) |> 
  rename(siteId = ID,
         timestamp = dttm) |> 
  mutate(dt_date = date(timestamp),
         dt_month = month(timestamp),
         dt_weekdays = weekdays(timestamp),
         dt_year = year(timestamp)) |> 
  select(siteId, dt_date, dt_year, dt_month, dt_weekdays, flow) -> dt23_sep_clean

df23_dt <- bind_rows(df23_dt, dt23_sep_clean)


## SCOOT 2022
df22 %>%  ## <- df is the raw SCOOT Data
  as_tibble() |>  
  mutate(timestamp = lubridate::as_datetime(dt),
         dt_date = date(timestamp),
         dt_month = month(timestamp),
         dt_weekdays = weekdays(timestamp),
         dt_year = year(timestamp))|> 
  filter(dt_month >= 8) |> 
  select(siteId, dt_date, dt_year, dt_month, dt_weekdays, flow) -> df22_dt


# Adding September 2023
dt22_sep |> 
  select(-year) |> 
  rename(siteId = ID,
         timestamp = dttm) |> 
  mutate(dt_date = date(timestamp),
         dt_month = month(timestamp),
         dt_weekdays = weekdays(timestamp),
         dt_year = year(timestamp)) |> 
  select(siteId, dt_date, dt_year, dt_month, dt_weekdays, flow) -> dt22_sep_clean

df22_dt <- bind_rows(df22_dt, dt22_sep_clean)



```


Let us take a look at the traffic monitors. Using the LEZ boundary `lez_shp`, we first draw a 50 metre buffer to include stations that are adjacent to the LEZ boundary. `mapview(lez_buff_50)` 

We will then clip the monitors according to the boundary. This throws us with 66 stations within the given boundary.

```{r}
# Wrap Buffer
lez_shp |> 
  st_transform(27700) |> 
  st_buffer(50) -> lez_buff_50

mapview(lez_buff_50)

# Filtering Monitor
lez_monitor <- st_intersection(monitor, lez_buff_50)

mapview(lez_monitor) +
  mapview(lez_buff_50)

```


Amongst the boundary, we choose four stations "GA1571_Q", "GA2401_D" for Hope Street, "GG2001_S", and "GA5371_C" for High Street.

```{r}
lez_monitor |> 
  filter(siteId %in% c("GA1571_Q", "GA2401_D",   ## Near GLA4
                       "GG2001_S", "GA5371_C" ## Near GHSR
  )) |> 
    mapview() +
  mapview(lez_buff_50)
```


We just want to double check the distance of traffic monitors from the air quality monitoring stations. Using the `openair` package, we sourced the Scottish air quality network and obtained two active stations, "GLA4"(Hope St) and "GHSR"(High St).

```{r}
lez_monitor |> 
  filter(siteId %in% c("GA2401_D",    ## Near GLA4
                       "GG2001_S" ## Near GHSR
  )) -> lez_monitor_for_ap

meta <- importMeta(source = "saqn", all = TRUE)

meta |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |> 
  st_intersection(lez_shp) |> 
  mapview()

meta |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |> 
  filter(code %in% c("GLA4", "GHSR"),
         variable == "NO2",
         end_date == 'ongoing') -> ap_glasgow ## Filter Glasgow
```

And the distance.
```{r}
## Distance 
st_distance(ap_glasgow |> st_transform(27700) |> filter(code == "GLA4"), 
            lez_monitor_for_ap |> filter(siteId == "GA2401_D"))

st_distance(ap_glasgow |> st_transform(27700) |> filter(code == "GHSR"), 
            lez_monitor_for_ap |> filter(siteId == "GG2001_S")) 
```


## SCOOT Data

Check `NA` counts.

```{r}

df22_dt |> 
  map( ~sum(is.na(.)))

df23_dt |> 
  #filter(dt_month == 8) |> 
  map( ~sum(is.na(.)))
#--> data between mid and end July are missing.
#--> 2023-07-12 13:00:00 to 2023-07-26 09:00:00

```


Visualisation.
```{r}
df22_dt %>% 
  filter(siteId %in% unique(lez_monitor_for_ap$siteId)) %>% 
  group_by(dt_date, siteId, dt_month) %>% 
  summarise(dailyflow = sum(flow)) %>% 
  ggplot(aes(dt_date, dailyflow, colour = siteId)) +
  geom_line() +
  theme(legend.position = "bottom") 

df23_dt %>% 
  filter(siteId %in% unique(lez_monitor_for_ap$siteId)) %>% 
  group_by(dt_date, siteId, dt_month) %>% 
  summarise(dailyflow = sum(flow)) %>% 
  ggplot(aes(dt_date, dailyflow, colour = siteId)) +
  geom_line() +
  theme(legend.position = "bottom") 
```


```{r}
traffic_tbl <- 
  bind_rows(df22_dt, df23_dt) |> 
  mutate(location = if_else(siteId == "GA1571_Q" | siteId == "GA2401_D", "Hope St", "High St"))

## Grouping Days of the Week into three categories
traffic_tbl %>% 
  filter(siteId %in% unique(lez_monitor_for_ap$siteId),
         dt_month != 7) %>% 
  mutate(week_group = case_when(dt_weekdays %in% c("Tuesday", "Wednesday", "Thursday") ~ "Core Weekdays",
                                dt_weekdays %in% c("Monday", "Friday") ~ "Other Weekdays",
                                dt_weekdays %in% c("Saturday", "Sunday") ~ "Weekends"
  )) %>% 
  group_by(siteId, location,week_group, dt_year, dt_date, dt_month) %>% 
  summarise(total_flow = sum(flow)) %>% 
  ungroup() -> df_for_analysis

df_for_analysis$dt_year <- factor(df_for_analysis$dt_year)

```

```{r}
library(ggh4x)

df_for_analysis |> 
  group_by(dt_date, siteId, location, dt_year, dt_month) %>% 
  summarise(dailyflow = sum(total_flow)) %>% 
  mutate(dt_month = if_else(dt_month == "8", "August", "September"),
         dt_month = factor(dt_month, levels = c("August", "September"))) |> 
  ggplot(aes(dt_date, dailyflow, colour = siteId)) +
  geom_smooth(linewidth = .3, linetype=9, alpha = .2) +
  geom_line() +
  labs(x = "", y = "Average daily flow") +
  facet_nested_wrap(vars(dt_month, dt_year), scales = "free_x") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),        
        text = element_text(size=13),
        axis.text.x = element_text(size = 8),
        legend.margin=margin(t=-20)) 

#ggsave("Figures/average daily traffic flow.jpg", width = 9, height = 5.5)

df_for_analysis |> 
  group_by(dt_date, siteId, location, dt_year, dt_month) %>% 
  summarise(dailyflow = sum(total_flow)) %>% 
  ggplot(aes(dt_date, dailyflow, colour = siteId)) +
  geom_smooth() +
  labs(x = "", y = "Average daily flow") +
  facet_wrap( ~dt_year , scales = "free_x") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        text = element_text(size=14),
        legend.margin=margin(t=-20)) 
```


```{r}
df_for_analysis |> 
  group_by(siteId, dt_year) |> 
  summarise(mean_daily_flow = mean(total_flow)) |>  
  pivot_wider(names_from = dt_year, values_from = mean_daily_flow)


df_for_analysis |> 
  group_by(siteId, week_group, dt_year) |> 
  summarise(mean_daily_flow = mean(total_flow)) |>  
  pivot_wider(names_from = dt_year, values_from = mean_daily_flow)


df_for_analysis |> 
  group_by(siteId, week_group, dt_year) |> 
  get_summary_stats(total_flow, type = "common")
```


```{r}
df_for_analysis |> 
  mutate(siteId = case_when(siteId == "GA2401_D" ~ "GA2401_D (Hope St)",
                            siteId == "GG2001_S" ~ "GG2001_S (High St)")) |> 
  gghistogram(x = "total_flow",
              add = "mean", 
              rug = TRUE,
              bins = 20,
              fill = "dt_year",
              color = "dt_year", 
              palette = c("#00AFBB", "#E7B800")) +
  labs(x = "Averaged Daily Flow", y = "Counts") +
  theme(legend.title = element_blank(),
        text = element_text(size=14)) 

## ggsave("Figures/traffic counts.jpg", width = 6, height = 3)

```


```{r}
df_for_analysis |> 
  mutate(siteId = case_when(siteId == "GA2401_D" ~ "GA2401_D (Hope St)",
                            siteId == "GG2001_S" ~ "GG2001_S (High St)")) |> 
  gghistogram(x = "total_flow",
            rug = TRUE,
            fill = "dt_year",
            bins = 15,
            color = "dt_year", 
            palette = c("#00AFBB", "#E7B800")) +
  ylim(0,20) +
  facet_wrap(~siteId, scales = "free") +
  labs(x = "Averaged Daily Flow", y = "Counts") +
  theme(legend.title = element_blank(),
        text = element_text(size=12)) 

#ggsave("Figures/traffic counts by station.jpg", width = 6.5, height = 4.5)


df_for_analysis |> 
  mutate(siteId = case_when(siteId == "GA2401_D" ~ "GA2401_D (Hope St)",
                            siteId == "GG2001_S" ~ "GG2001_S (High St)")) |> 
  ggboxplot(x = "dt_year", y = "total_flow", color = "dt_year") +
  facet_wrap(~siteId, scales = "free") +
  stat_summary(fun.y = mean, geom="point",colour="darkred", size=3) +
  labs(x = "", y = "Averaged Daily Flow") +
  theme(legend.title = element_blank(),
        text = element_text(size=12),
        axis.text.x = element_text(size=10)) 

#ggsave("traffic counts boxplot.jpg", width = 6, height = 4.5)

# Calculate means manually
mean_data <- df_for_analysis %>%
  group_by(week_group, dt_year, siteId) %>%
  summarize(mean_total_flow = mean(total_flow), .groups = 'drop')


# Calculate means manually
mean_data <- df_for_analysis %>%
    mutate(siteId = case_when(siteId == "GA2401_D" ~ "GA2401_D (Hope St)",
                            siteId == "GG2001_S" ~ "GG2001_S (High St)"),
         week_group = factor(week_group)) |> 
  group_by(week_group, dt_year, siteId) %>%
  summarize(mean_total_flow = mean(total_flow), .groups = 'drop')


df_for_analysis |> 
  mutate(siteId = case_when(siteId == "GA2401_D" ~ "GA2401_D (Hope St)",
                            siteId == "GG2001_S" ~ "GG2001_S (High St)"),
         week_group = factor(week_group)) |> 
ggplot(aes(x = week_group, y = total_flow, fill = dt_year)) +
  geom_boxplot(aes(group=interaction(dt_year, week_group)), position=position_dodge(0.75)) +
  geom_point(data = mean_data, aes(x = week_group, y = mean_total_flow, group = interaction(dt_year, week_group)),
             position=position_dodge(0.75), size=2, color="darkred") +
  facet_wrap(~siteId, scales = "free_y") +
  labs(x = "", y = "Averaged Daily Flow") +
  theme_pubr() +
  theme(legend.title = element_blank(),
        text = element_text(size=12),
        axis.text.x = element_text(size=10))

#ggsave("traffic counts boxplot weekgroup.jpg", width = 8.5, height = 4.5)

```



